---
title: "Tutorial"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE)
```

## Welcome
### Outline


Welcome to Tutorial 2. In this tutorial we will show you how to:

1. take a random sample from a finite population
2. take a random sample from an infinite population
3. describe sample data with the mean and 95% confidence interval


## Part 1: Sampling - finite populations
### How do I take a random sample?

There are 2 types of populations from which we may wish to take a random sample; finite and infinite populations.
Finite populations are made of discrete objects like households, people, farms etc. Infinite populations are continuous, such as soil, water in a stream. 

### Finite populations

Randomly sampling from a finite population can easily be performed in R.
Imagine a hypothetical situation where we know there are 100 farms in a catchment and we wish to survey 10 of them about their management practices.
First, we number the farms from 1 to 100:

```{r}
farms <- 1:100
```

The line above creates an object which is a vector with numbers from 1 to 100 going up in increments of 1. Each of these indicates a different farm. We now wish to randomly choose 10 farms to survey.

```{r}
set.seed(42) # this makes functions that depend on random number generators reproducible
sites <- sample(x = farms, size = 10)
sites
```


```{r sample1, exercise = TRUE}
# playground

```


1. The `sample()` function has 2 main arguments (inputs), a vector as its first argument which defines the members of the population, and a number indicating how many to sample. 
2. We assign `sites` to the output of `sample()`. This output is *also* a vector of numbers.
3. Due to the randomness of the `sample()` function, `sites` will produce different outputs whenever you re-run the code, but `set.seed()` can "fix" the outputs to be identical to anyone who uses the same seed value.

**The advantage of `set.seed()` is that we can reproduce random sampling by code, which makes it much easier to design experiments.**

::: {.alert .alert-warning}
**Exercise**
Try deleting, commenting out (`#`), or changing the value of `set.seed()`, to see what it does to the output. How would you use `set.seed()` in an experimental design that prevents bias?

:::



```{r setseed, exercise=TRUE}
# first, define your `farms` object, and assign numbers 1 to 100 to it
# then, sample the farms data

```




## Part 2: Sampling - infinite populations
### How do I sample a spatial location?

Here, we are mainly considering infinite populations and we are interested in estimating the mean of a measurement, spatially, across a study area. In this case we need to randomly choose a spatial location which is defined by a spatial coordinate, x and y. 

Assuming a *hypothetical* study site 100m wide and 100m long where we wish to choose 10 random locations, then we may use the technique below to randomly select 10 locations based on x and y coordinates:

```{r}
set.seed(101)
xcoord <- runif(n = 10, min = 0, max = 100)
ycoord <- runif(n = 10, min = 0, max = 100)

sites_xy <- data.frame(xcoord, ycoord)

```


```{r spatialSample, exercise=TRUE}
# playground

```


### Using `runif()`

The `runif()` function generates random numbers with a uniform distribution. The 3 main arguments (inputs) are:

1. `x`: the number of samples, 
2. `min`: the minimum value we want in our random numbers
3. `max`: the minimum value we want in our random numbers 

How did we know this information? Use `?runif` and try to see where you can find similar documentation in the "Help" pane that appears.

::: {.alert .alert-warning}
**Exercise**
How would you randomly sample 12 locations from a study site 50m wide and 50m long? Write your code below as practice.
:::

```{r sample_runif, exercise=TRUE}
# don't forget to use `set.seed()`

```

### Plotting

In our original example we choose 0 and 100 as our study area is 100 m wide and 100 m long. The random numbers are generated twice; once for the x-coordinates and again for the y-coordinates. We can now plot our sample locations.

```{r}
# base R plot
plot(sites_xy, xlab = "x-coordinates (m)", ylab = "y-coordinates (m)")
```

One issue is that not all study areas are rectangles. Rather they are quite often irregular. This artificial dataset is just to illustrate the **concept** rather than be practically useful. The analysis of spatial data is not in the scope of this unit.

## Part 2 - Simple random design

### How do we summarise a simple random sample?

In the case of a simple random sampling scheme we can manually calculate the mean and 95% CI to describe the sample data. You will be doing this during your lab.

Alternatively, we could just use the `t.test()` function to estimate the mean and the associated 95% CI while ignoring the rest of the output, that is, *ignore the p-value and the null hypothesis*. **This works because the samples were sampled randomly, therefore the sampling distribution should follow a t-distribution.**
Below is an example from the soil carbon dataset in the lecture.

```{r}
soil <- c(48, 56, 90, 78, 86, 71, 42)
t.test(x = soil)
```

```{r soil, exercise=TRUE}
# playground


```

## Finish

This concludes this week's tutorial.