---
title: "Lab 2 - Surveying and Monitoring"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

## Learning outcomes
At the end of this practical students should be able to:

- explain conceptually the benefits of re-sampling (or not) the same units for monitoring studies.
- use R to create rudimentary sampling designs;
- estimate means and associated CIs for simple random designs;
- estimate means and associated CIs for stratified random designs;
- use R for estimating the change in means (and associated CIs) for monitoring schemes when the sample units are (i) resampled (ii) not resampled.

## Topic 1 - Stratification

### To stratify or not to stratify?

Karunaratne et al. (2012)^[S.B. Karunaratne, T. Bishop, I. Odeh, J. Baldock, B. Marchant, Estimating change in soil organic carbon using legacy data as the baseline: Issues, approaches and lessons to learn, Soil Research, 52 (2014) 349-365.] measured soil carbon (%) in the Cox’s creek catchment in northern NSW using a stratified random sampling scheme. The results are summarised below. 

```{r setupA, echo=FALSE}
carbon <-
  data.frame(
    use = c(
      "Forest", "Dryland Cropping", "Pasture-Vertosol", "Pasture-Other",
      "Irrigated"
    ),
    n = c(9, 14, 14, 2, 5),
    mean = c(0.65, 0.96, 1.06, 1.31, 0.78),
    variance = c(1.4, 0.4, 0.8, 0.4, 0.8),
    percent_area = c(20, 35, 35, 6, 4)
  )

overall <-
  data.frame(
    use = "Overall", n = 44, mean = 0.92, variance = 1,
    percent_area = 100
  )

knitr::kable(rbind(carbon, overall))
```

Answer the next three questions using the data. Since we were not given the original data, but a summary table, we need to back-calculate our summaries of the data. 

::: {.alert .alert-info}
**Note** Your **Week 2 lecture notes** contain R code that can help you with your calculations of the 95% CI for stratified random sampling and simple random sampling.
:::

### Exercise 1.1

Estimate the mean and associated 95% CI assuming **stratified** random sampling using the `carbon` dataset.

Steps:

1. Calculate the weight of each stratum to calculate weighted mean per stratum
2. Calculate (weighted variance of the mean)^2^, and derive the standard error values
3. Determine 95% CI via the equation below:

$$95\% CI = \bar y \pm t^{0.025}_{n-L} \times se(\bar y)$$

where $L$ = number of strata

You can access the results by evaluating the `carbon` object in the code:

```{r echo=TRUE, results=FALSE}
carbon
```

```{r 1.1}
# write your code here

```

### Exercise 1.2

Estimate the mean and associated 95% CI assuming **simple** random sampling using the `overall` dataset.

```{r 1.2}
overall
```

```{r 1b}
# write your code here

```

### Exercise 1.3

Would you recommend stratified random sampling for future surveys? Provide evidence from the results of questions 2 and 3.


## Topic 2 - Monitoring

### Should we return to the same locations?

In this exercise we will explore the impact that resampling the locations has on the precision (width of the 95% CI) with which we estimate the change in mean between 2 surveys. We will also show the equivalence of this to 2-sample t-tests.

The key equation for estimating the variance of the change in mean is:

$$Var (\Delta\bar{y}) = 
  Var (\Delta\bar{y}_2) + 
  Var (\Delta\bar{y}_1) - 
  2 \times Cov(y_1, y_2) $$

where:

- $Var (\Delta\bar{y})$ is the variance of the change in mean between 2 surveys;
- $Var (\Delta\bar{y}_2)$ is the variance of the mean for the 1st survey (baseline);
- $Var (\Delta\bar{y}_1)$ is the variance of the mean for the 2nd survey (repeat);
- $Cov(y_1, y_2)$ is the covariance between the means of the 2 surveys.

When we resample the same locations we include the covariance term which describes the relationship between the observations in the 2 surveys. If resample at **different locations** we assume the **covariance is equal to 0**.

In this exercise we will consider a study where kangaroos (number/km^2^) were counted in a woodland in an initial survey, and then 2 years later.
The data is below.

```{r, echo=FALSE}
baseline <- c(4, 2, 6, 1, 3)
rerun <- c(12, 11, 13, 14, 9)

kangaroos <- data.frame(baseline, rerun)
str(kangaroos)
```

### Exercise 2.1

Intuitively, what would be the impact on the precison with which we estimate the mean when 

(a) we resample the same locations in the repeat survey? 
(b) we sample at different locations in the repeat survey?

### Exercise 2.2

Assuming we resampled at **different locations** to the baseline survey, use R to estimate the 95% CI for the change in mean between the 2 surveys. The equation is:

$$95\%\ CI=\Delta\left(\bar{y}\right)\pm t_{df}^{0.025}\times\sqrt{Var\left(\Delta\bar{y}\right)}$$


You will need to use the R calculations above to calculate ${Var}\left(\Delta\left(\bar{y}_1\right)\right)$, $Var\left(\Delta\left(\bar{y}_2\right)\right)$, and $Cov\left(\bar{y}_1,\bar{y}_2\right)$.

As a starting point R script for calculating the covariance and variance for the **observations** is below, as is the code to calculate the t critical value ($t_{df}^{0.025}$). When we sample at different locations between surveys:
$$df = n_{1}+ n_{2} - 2 = 5 + 5 - 2 = 8$$

```{r echo=TRUE}
t0 <- kangaroos$baseline
t2 <- kangaroos$rerun
# covariance
cov(t0, t2)

# variances
var(t0)
var(t2)

# t-critical
tcrit1 <- qt(0.975, 8)
```

### Exercise 2.3

Assuming we resampled the **same locations** as the baseline survey, use R to estimate the 95% CI for the change in mean between the 2 surveys. Note: when we sample at same locations between surveys the *df* = number of paired sites - 1 = 5 - 1 = 4. 

Hint: the following formulae are used:

$$\Delta \bar y = \bar y_2 - \bar y_1$$
$$Var(\Delta \bar y) = Var( \bar y_2) + Var( \bar y_1) - 2 \times Cov(\bar y_1, \bar y_2)$$



### Exercise 2.4

For the scenarios in Exercise 2.2 and Exercise 2.3, was there a significant change in the mean number of kangaroos between the study period?

### Exercise 2.5

In the future which would you do? Re-sample the same sites or visit new sites in a monitoring program? If it were soil how would your thinking change? To help you answer the question compare the results with the soil monitoring example in the Lecture.

### Exercise 2.6

The scenario when we **resample at different locations** between surveys is actually the equivalent to a **2-sample t-test**, and the scenario when we **resample the same locations** is equivalent to a **paired t-test**. Check your answers in R by trying to perform both of these using the `t.test` function. Look at the help command, for example type `?t.test` to see what you can work out. Assume the variances are equal for the 2-sample t-test (use the argument `var.equal = TRUE`for this).

### Exercise 2.7

Compare the 95% CI for each scenario you calculated manually in Exercises 3 and 4. Are they same?